sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-08-08 doc/sdmTMB-PeerJ.Rmd
Reformat

---
title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
author: "|
  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^3$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^4$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    bookdown::word_document2:
    # bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
      keep_tex: true
bibliography: refs.bib
link-citations: yes
csl: ecography.csl
header-includes:
  \usepackage[document]{ragged2e}
  \usepackage{setspace}\doublespacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \setlength{\parskip}{1.0\baselineskip}
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{lscape}
  \widowpenalty10000
  \clubpenalty10000
---

@@

\clearpage

\linenumbers

\setcounter{page}{1}

# sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

# Abstract

Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. 
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. 
Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) based approach. 
Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. 
We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru). 
Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. 
Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. 
We anticipate that its application to the full spectrum of problems involving geostatistical data will prove equally fruitful.

Keywords:
Gaussian Markov random fields (GMRF),
SPDE,
species distribution modelling,
R package,
spatio-temporal or spatial-temporal,
Template Model Builder

\clearpage

@@

\onehalfspacing

@@

\doublespacing

>>>>

---
title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
preprint: false
author:
  - name: Sean C. Anderson
    affiliation: 1
    corresponding: true
    email: sean.anderson@dfo-mpo.gc.ca
  - name: Eric J. Ward
    affiliation: 2
  - name: Philina A. English
    affiliation: 1
  - name: Lewis A. K. Barnett
    affiliation: 3
affiliation:
  - code: 1
    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - code: 2
    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
  - code: 3
    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
output:
    bookdown::pdf_book:
      template: template-peerj.tex
      pandoc_args: "--highlight=haddock"
      toc: false
      number_sections: false
      keep_tex: true
abstract: >
  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions.  However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement.  Therefore, spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways. Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE- (stochastic partial differential equation) based approach.  Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible.  We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software. The case studies illustrate sdmTMB’s unmatched flexibility and usability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru).  Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation.  Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru. sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling.  We anticipate that its application to the full spectrum of problems involving geostatistical data will prove equally fruitful.
bibliography: refs.bib
link-citations: yes
csl: peerj.csl
header-includes:
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \widowpenalty10000
  \clubpenalty10000
---

@@

<!-- \linenumbers -->

<!-- Keywords: -->
<!-- Gaussian Markov random fields (GMRF), -->
<!-- SPDE, -->
<!-- species distribution modelling, -->
<!-- R package, -->
<!-- spatio-temporal or spatial-temporal, -->
<!-- Template Model Builder -->

@@

@@
